A spatial analysis of the environmental correlates of COVID-19 incidence in Spain

### 2020-04-18 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
cronologia

@@

Our dataset includes information about the daily number of cases of COVID-19 reported in Spain \footnote{cases tested} at the provincial level (NUTIII in eurostat terminology) from March 13th to April 11th (30 days and 50 spatial units). The Spanish National Government publishes periodic updates at the regional level (NUTII) and the information is also released at the provincial level as part of a collaborative project. This information is compiled from several sources, mainly the official web pages of the Spanish regional goverments. In addition, we consider two sets of explanatory variables. The first one, and the focus of this research, is set of two environmental variables, namely temperature and humidity. The second set provides some relevant controls for multivariate analysis, and refers to economic and demographic attributes of the province. Table \ref{tab:descriptive-statistics} shows the descriptive statistics and the provencance of the data.

@@

Seemingly Unrelated Regressions are used in the context of...

@@

>>>>

For a better interpretation of the results of our analysis is necessary to know the chronology of the epidemic outbreak in Spain. This subsection presents a brief time line of COVID-19.

The first reported case of COVID-19 in Spain was on January 31th, were a German tourist in the 
Canary Islands was positively tested, but the first not imported case was diagnostic on February 27th in Sevilla province (Andalusia). The epidemic was growing slowly at the beginning but in March 11th the World Health Organization (WHO) declared officially the pandemic. This declaration was a break point and on March 13th the Spanish National Government declare the "Emergency Status". At this date, Spain have been declared 4209 cases of COVID-19 (mostly of cases, 1990, in Madrid). Only essential services (e.g. food, health) and some economic subsector of industry and construction could to maintain activity. On March 17th Spain closes its lands borders and only allows entry to nationals and residents. From March 30th to April 12th (April 10th-12th Easter) lockdown of the Spanish people was hardened and only essential activities were allowed.

@@

Our dataset includes information about the daily number of cases of COVID-19 reported in Spain \footnote{cases positively tested. Exclude sintomatic cases diagnostic by doctor with-out Reacción en Cadena de la Polimersa (PCR) test} at the provincial level (NUTIII in eurostat terminology) from March 13th to April 11th (30 days and 50 spatial units). The Spanish National Government publishes periodic updates at the regional level (NUTII) and the information is also released at the provincial level as part of a collaborative project. This information is compiled from several sources, mainly the official web pages of the Spanish regional goverments. In addition, we consider two sets of explanatory variables. The first one, and the focus of this research, is set of two environmental variables, namely temperature and humidity. The second set provides some relevant controls for multivariate analysis, and refers to economic and demographic attributes of the province. Table \ref{tab:descriptive-statistics} shows the descriptive statistics and the provencance of the data.

@@

The spatial Seemingly Unrelated Regressions (SUR) methodology has been a popular method to estimate are used in the context of...

The Seemingly Unrelated Regression equations model (SUR from now on) is a multivariate econometric formulation used in different fields when the data set is organized as cross-sections for different time periods. The basis of this approach is well-known since the initial works of @zellner1962 and now is a popular methodology included in several Econometrics textbook (e.g. @greene2003). To our knowledge, @Anselin1988spatial was the first text that introduced the term spatial SUR in reference to a model made of ‘an equation for each time period, which is estimated for a cross section of spatial units’ (p. 141). From this milestone, a huge researches used this methodology in spatial framework. @rey1999,@fingleton2007multi, @lauridsen2010, @legallo2006, @lopez2017spatial are well-known examples in this line.

@@

Another alternative methodologies could be use. By example, a dynamic spatial panel methodology with fixed spatial an temporal effects (e.g. Elhorst 2014, Cap. 4), but those models are more dont take account correlation between errors. Therefore, a spatial SUR approach is more reasonable for our purpose.

### 2020-04-18 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.tex
cronologia

  \begin{abstract}
  Spreading with astonishing speed, the novel SARS-CoV2 has swept the
  globe, causing enormous stress to health systems and prompting social
  distance guidelines and mandates to arrest its progress. While there is
  encouraging evidence that early public health interventions have slowed
  the spread of the virus, this has come at a high cost as the global
  economy is brought to its knees. How and when to ease restrictions to
  movement hinges in part on the question whether SARS-CoV2 will display
  seasonality associated with variations in temperature and humidity. In
  this research, we address this question by means of a spatial analysis
  of the incidence of COVID-19 in the provinces in Spain. Use of a spatial
  Seemingly Unrelated Regressions (SUR) approach allows us to model the
  incidence of reported cases of the disease per 100,000 population, as a
  function of temperature and humidity, while controlling for GDP per
  capita, population density, percentage of older adults in the
  population, and presence of mass transit systems. Our results indicate
  that incidence of the disease is lower at higher temperatures. The
  evidence with respect to humidity is more mixed, with coefficients for
  this variable that are significant in only some equations. Our control
  variables also yield interesting insights, as population density and
  percentage of older adults display negative associations with incidence
  of COVID-19, whereas the presence of mass transit systems in the
  province is associated with a greater incidence of the disease.
  \end{abstract}

@@

Some of what is thought about the possible seasonality of COVID-19 is
based on analogies to the patterns of other known respiratory viruses.
However, de Ángel Solá et al.~(2020) note that ``not all seasonal
respiratory viruses experience the same spatiotemporal patterns''
(section 4). This urges caution when extrapolating from known viruses,
and indeed, the evidence in this respect is inconclusive. At a global
scale, whereas de Ángel Solá et al.~(2020) see less risk in the Caribean
Basin, Coelho et al.~(Coelho et al., 2020) warn that at least during the
exponential phase, expansion of the virus is not driven by climate.
Similarly, whereas Araujo and Naimi (2020) argue that spread of
SARS-CoV2 will likely be constrained by climate, Harbert et al.~(2020)
remain unconvinced that spatial modelling can currently discriminate the
distribution of the disease on the basis of climate, at least in the
United States. To further complicate matters, much of the relevant work
has yet to be peer-reviewed and therefore is open to change (see for
example the challenge of Harbert et al. (2020) to Araujo and Naimi
(2020)). In the United States, the National Academy of Sciences,
Engineering, and Medicine was engaged by the Office of the Executive for
guidance on this matter (see National Academies of Sciences, Engineering
and Medicine, 2020). Their conclusion summarizes the situation well (see
p.~6): ``Some limited data support a potential waning of cases in warmer
and more humid seasons, yet none are without major
limitations\ldots Additional studies as the SARS-CoV-2 pandemic unfolds
could shed more light on the effects of climate on transmission.''

With the above considerations in mind, our objective with this paper is
to contribute to the knowledge basis regarding the spread of COVID-19
and the influence of environmental factors, particularly temperature and
humidity. Hence, here we report results from a spatial analysis of
incidence of COVID-19 in fifty provinces in Spain. Spain is one of the
countries hardest hit by the virus, and enacted lockdown measures on
March 16, 2020, in response to a rapidly growing outbreak of COVID-19.
We combine data on reported cases of the disease with metereological
information, to create a spatio-temporal dataset covering a period of 30
days. We then join this dataset with provincial-level economic and
demographic information to act as controls to our key environmental
variables. These data are analyzed using a spatial Seemingly Unrelated
Regressions (SUR) approach, which allows us to account for spatial
autocorrelation.

The results provide evidence of the effect of temperature on the
incidence of COVID-19, as \textbf{NOTE ABOUT THE MAGNITUDE OF THE
EFFECT}. The evidence concerning humidity is more mixed: while the
direction of the effect is negative, as anticipated, the coefficients
for this variable are only significant in some of equations in the
system. Our control variables also provide some intriguing insights. The
results of this analysis provide support to the hypothesis of
seasonality of the novel SARS-CoV2, and should be of interest to public
health officials and policy makers wrestling with the question of the
trajectory of the pandemic.

@@

\begin{quote}
\begin{quote}
Include link to repository
\end{quote}
\end{quote}

@@

@@

Our dataset includes information about the daily number of cases of
COVID-19 reported in Spain \footnote{cases tested} at the provincial
level (NUTIII in eurostat terminology) from March 13th to April 11th (30
days and 50 spatial units). The Spanish National Government publishes
periodic updates at the regional level (NUTII) and the information is
also released at the provincial level as part of a collaborative
project. This information is compiled from several sources, mainly the
official web pages of the Spanish regional goverments. In addition, we
consider two sets of explanatory variables. The first one, and the focus
of this research, is set of two environmental variables, namely
temperature and humidity. The second set provides some relevant controls
for multivariate analysis, and refers to economic and demographic
attributes of the province. Table \ref{tab:descriptive-statistics} shows
the descriptive statistics and the provencance of the data.

@@

Seemingly Unrelated Regressions are used in the context of\ldots{}

@@

\noindent where\ldots.

@@

>>>>

  \begin{abstract}
  Spreading with astonishing speed, the novel SARS-CoV2 has swept the
  globe, causing enormous stress to health systems and prompting social
  distance guidelines and mandates to arrest its progress. While there is
  encouraging evidence that early public health interventions have slowed
  the spread of the virus, this has come at a high cost as the global
  economy is brought to its knees. How and when to ease restrictions to
  movement hinges in part on the question whether SARS-CoV2 will display
  seasonality associated with variations in temperature and humidity. In
  this research, we address this question by means of a spatial analysis
  of the incidence of COVID-19 in the provinces in Spain. Use of a spatial
  Seemingly Unrelated Regressions (SUR) approach allows us to model the
  incidence of reported cases of the disease per 100,000 population, as a
  function of temperature and humidity, while controlling for GDP per
  capita, population density, percentage of older adults in the
  population, and presence of mass transit systems. An interesting aspect
  of the spatial SUR approach is that it models incidence as a contagion
  process. Our results indicate that incidence of the disease is lower at
  higher temperatures. The evidence with respect to humidity is more
  mixed, with coefficients for this variable that are significant in only
  some equations. Our control variables also yield interesting insights,
  as population density and percentage of older adults display negative
  associations with incidence of COVID-19, whereas the presence of mass
  transit systems in the province is associated with a greater incidence
  of the disease.\\
  \end{abstract}

@@

Some of what is thought about the possible seasonality of COVID-19 is
based on analogies to the patterns of other known respiratory viruses.
However, de Ángel Solá et al.~(2020) note that ``not all seasonal
respiratory viruses experience the same spatiotemporal patterns''
(section 4). This urges caution when extrapolating from known viruses,
and indeed, the evidence in this respect is inconclusive. At a global
scale, whereas de Ángel Solá et al.~(2020) see less risk in the Caribean
Basin, Coelho et al.~(Coelho et al., 2020) warn that at least during the
exponential phase, expansion of the virus is not driven by climate.
Similarly, whereas Araujo and Naimi (2020) argue that spread of
SARS-CoV2 will likely be constrained by climate, Harbert et al.~(2020)
remain unconvinced that spatial modelling can currently discriminate the
distribution of the disease on the basis of climate, at least in the
United States. To further complicate matters, much of the relevant work
has yet to be peer-reviewed and therefore is open to change (see for
example the challenge of Harbert et al. (2020) to Araujo and Naimi
(2020)). In the United States, the National Academy of Sciences,
Engineering, and Medicine was engaged by the Office of the Executive for
guidance on this matter (see National Academies of Sciences, Engineering
and Medicine, 2020). Their conclusion summarizes the situation well (see
p.~6): ``Some limited data support a potential waning of cases in warmer
and more humid seasons, yet none are without major
limitations\ldots{}Additional studies as the SARS-CoV-2 pandemic unfolds
could shed more light on the effects of climate on transmission.''

With the above considerations in mind, our objective with this paper is
to contribute to the knowledge basis regarding the spread of COVID-19
and the influence of environmental factors, particularly temperature and
humidity. Hence, here we report results from a spatial analysis of
incidence of COVID-19 in fifty provinces in Spain. Spain is one of the
countries hardest hit by the virus, and enacted lockdown measures on
March 16, 2020, in response to a rapidly growing outbreak of COVID-19.
We combine data on reported cases of the disease with metereological
information, to create a spatio-temporal dataset covering a period of 30
days. We then join this dataset with provincial-level economic and
demographic information to act as controls to our key environmental
variables. These data are analyzed using a spatial Seemingly Unrelated
Regressions (SUR) approach, which allows us to model incidence of
COVID-19 as a spatial contagion process.

The results provide evidence of the effect of temperature on the
incidence of COVID-19. The evidence concerning humidity is more mixed:
while the direction of the effect is negative, as anticipated, the
coefficients for this variable are only significant in some of equations
in the system. Our control variables also provide some intriguing
insights. The results of this analysis provide support to the hypothesis
of seasonality of the novel SARS-CoV2, and should be of interest to
public health officials and policy makers wrestling with the question of
the trajectory of the pandemic.

@@

https://github.com/paezha/covid19-environmental-correlates

@@

For a better interpretation of the results of our analysis is necessary
to know the chronology of the epidemic outbreak in Spain. This
subsection presents a brief time line of COVID-19.

The first reported case of COVID-19 in Spain was on January 31th, were a
German tourist in the Canary Islands was positively tested, but the
first not imported case was diagnostic on February 27th in Sevilla
province (Andalusia). The epidemic was growing slowly at the beginning
but in March 11th the World Health Organization (WHO) declared
officially the pandemic. This declaration was a break point and on March
13th the Spanish National Government declare the ``Emergency Status''.
At this date, Spain have been declared 4209 cases of COVID-19 (mostly of
cases, 1990, in Madrid). Only essential services (e.g.~food, health) and
some economic subsector of industry and construction could to maintain
activity. On March 17th Spain closes its lands borders and only allows
entry to nationals and residents. From March 30th to April 12th (April
10th-12th Easter) lockdown of the Spanish people was hardened and only
essential activities were allowed.

@@

Our dataset includes information about the daily number of cases of
COVID-19 reported in Spain
\footnote{cases positively tested. Exclude sintomatic cases diagnostic by doctor with-out Reacción en Cadena de la Polimersa (PCR) test}
at the provincial level (NUTIII in eurostat terminology) from March 13th
to April 11th (30 days and 50 spatial units). The Spanish National
Government publishes periodic updates at the regional level (NUTII) and
the information is also released at the provincial level as part of a
collaborative project. This information is compiled from several
sources, mainly the official web pages of the Spanish regional
goverments. In addition, we consider two sets of explanatory variables.
The first one, and the focus of this research, is set of two
environmental variables, namely temperature and humidity. The second set
provides some relevant controls for multivariate analysis, and refers to
economic and demographic attributes of the province. Table
\ref{tab:descriptive-statistics} shows the descriptive statistics and
the provencance of the data.

\caption{\label{tab:descriptive-statistics}\label{tab:descriptive-statistics}Descriptive statistics}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{l>{\raggedright\arraybackslash}p{15em}rrrr>{\raggedright\arraybackslash}p{5em}}
\toprule
Variable & Note & Min & Mean & Max & SD & Source\\
\midrule
\rowcolor{gray!6}  COVID-19 Incidence & Incidence in reported cases of SARS-19 per 100,000 people & 0.38 & 153.61 & 1149.36 & 186.23 & Montera34\\
GDPpc & GDP per capita in €1,000s & 16.67 & 22.51 & 36.00 & 4.77 & INE\\
\rowcolor{gray!6}  Older & Percentage of people aged 65 and older in the province & 15.16 & 21.03 & 31.36 & 3.95 & INE\\
Population Density & Population density in the province in people per sq.km & 8.60 & 140.04 & 829.76 & 181.25 & INE\\
\rowcolor{gray!6}  Mean Temperature & Mean temperature in province by date, in C & 1.00 & 12.18 & 23.20 & 3.67 & AEMET\\
\addlinespace
Humidity & Relative humidity in province by date & 2.00 & 77.82 & 100.00 & 10.37 & MAPA\\
\bottomrule
\multicolumn{7}{l}{\textit{Note: }}\\
\multicolumn{7}{l}{Montera34: https://code.montera34.com:4443/numeroteca/covid19/-/blob/master/README\_providencialdata19.md}\\
\multicolumn{7}{l}{INE (Instituto Nacional de Estadistica): https://www.ine.es/}\\
\multicolumn{7}{l}{AEMET (Agencia Estatal de Meteorologia): http://eportal.mapa.gob.es}\\
\multicolumn{7}{l}{MAPA (Ministerio de Agricultura, Pesca y Alimentacion): http://eportal.mapa.gob.es}\\
\end{tabular}}
\end{table}

@@

The spatial Seemingly Unrelated Regressions (SUR) methodology has been a
popular method to estimate are used in the context of\ldots{}

The Seemingly Unrelated Regression equations model (SUR from now on) is
a multivariate econometric formulation used in different fields when the
data set is organized as cross-sections for different time periods. The
basis of this approach is well-known since the initial works of
({\textbf{???}}) and now is a popular methodology included in several
Econometrics textbook (e.g.~({\textbf{???}})). To our knowledge, Anselin
(1988) was the first text that introduced the term spatial SUR in
reference to a model made of `an equation for each time period, which is
estimated for a cross section of spatial units' (p.~141). From this
milestone, a huge researches used this methodology in spatial framework.
({\textbf{???}}),({\textbf{???}}), ({\textbf{???}}), ({\textbf{???}}),
({\textbf{???}}) are well-known examples in this line.

@@

\noindent where\ldots{}.

@@

Another alternative methodologies could be use. By example, a dynamic
spatial panel methodology with fixed spatial an temporal effects
(e.g.~Elhorst 2014, Cap. 4), but those models are more dont take account
correlation between errors. Therefore, a spatial SUR approach is more
reasonable for our purpose.
